---
title: "AlderIPM-Sim: Complete Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AlderIPM-Sim: Complete Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Introduction

The **alder-ipm-sim** package provides tools for simulating, fitting, and
analysing the Alnus glutinosa -- beetle -- parasitoid -- bird ecoepidemic
system. This vignette walks through a complete workflow from parameter
exploration to management recommendations.

## 1. Parameter Exploration

```{r params}
library(alderIPMSim)

# View default parameters
params <- default_params()
print(params)

# Explore the full registry with metadata
reg <- param_registry()
cat("Parasitoid attack rate (beta):", reg$beta$default, reg$beta$unit, "\n")
cat("Description:", reg$beta$description, "\n")

# Pretty-print mortality parameters
print_params(category = "mortality")
```

## 2. Multi-Year Simulation

Simulate 50 years of ecosystem dynamics starting from typical initial
conditions.

```{r simulation}
p <- as.list(params)
p$u_C <- 0
p$u_P <- 0

traj <- simulate(p, A0 = 0.8, F0 = 0.1, K0 = 1.712, D0 = 0, n_years = 50)

par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(traj$year, traj$A, type = "l", col = "firebrick",
     xlab = "Year", ylab = "Beetle density (A)", main = "Beetle Population")
plot(traj$year, traj$F, type = "l", col = "steelblue",
     xlab = "Year", ylab = "Parasitoid density (F)", main = "Parasitoid Population")
plot(traj$year, traj$K, type = "l", col = "forestgreen",
     xlab = "Year", ylab = "Carrying capacity (K)", main = "Canopy Health")
plot(traj$year, traj$D, type = "l", col = "darkorange",
     xlab = "Year", ylab = "Defoliation (D)", main = "Defoliation")
```

## 3. Equilibrium Analysis

Compute the parasitoid reproduction number R_P.

```{r rp}
R_P <- compute_RP(p)
cat("Parasitoid reproduction number R_P =", round(R_P, 3), "\n")
if (R_P > 1) {
  cat("R_P > 1: parasitoid can persist -- biological control is viable.\n")
} else {
  cat("R_P < 1: parasitoid cannot invade -- augmentation needed.\n")
}
```

## 4. Early Warning Signal Detection

Use the simulated beetle density trajectory to demonstrate early warning
signal detection.

```{r ews}
beetle_ts <- traj$A

ews <- compute_ews(beetle_ts, window = 10)
head(ews)

# Kendall trend test on variance
kt <- kendall_trend(ews$variance)
cat("Kendall tau for variance:", round(kt$tau, 3),
    " p-value:", round(kt$p_value, 4), "\n")
```

## 5. Loading Example Data

The package ships with example CSV datasets in `inst/extdata/`.

```{r extdata}
stable_file <- system.file("extdata", "example_stable_coexistence.csv",
                           package = "alder-ipm-sim")
stable_data <- read.csv(stable_file)
head(stable_data)

# Prepare for fitting
prepared <- prepare_data(stable_data, time_col = "year")
cat("Prepared", prepared$n_obs, "observations with states:",
    paste(names(prepared$obs), collapse = ", "), "\n")
```

## 6. Data Fitting

Fit model parameters to observed data.

```{r fitting, eval = FALSE}
result <- fit_model(prepared, fit_params = c("beta", "phi"))
cat("R-squared:", round(result$r_squared, 4), "\n")
cat("Fitted beta:", round(result$fitted_params[["beta"]], 5), "\n")
cat("Fitted phi:", round(result$fitted_params[["phi"]], 5), "\n")
```

## 7. Management Strategy Comparison

Compare three integrated pest management strategies.

```{r control, eval = FALSE}
ic <- c(A = 0.8, F = 0.1, K = 1.712, D = 0)
comp <- compare_strategies(p, initial_state = ic, n_years = 30)
print(comp$comparison)
cat("\n", comp$interpretation, "\n")
```

## 8. Interactive Dashboard

Launch the Shiny dashboard for interactive exploration:

```{r shiny, eval = FALSE}
run_app()
```

## Summary

This vignette demonstrated the core AlderIPM-Sim workflow:

1. **Parameter exploration** -- inspect defaults and metadata
2. **Simulation** -- run multi-year ecosystem projections
3. **Equilibrium analysis** -- compute R_P and stability
4. **Early warnings** -- detect critical slowing down
5. **Data fitting** -- estimate parameters from field data
6. **Control optimisation** -- compare management strategies
7. **Interactive dashboard** -- explore interactively via Shiny
